---
title: "Montgomery Integrative DESeq2 Pipeline"
author: "Dr. Taiowa Montgomery and Spencer Kuhn"
date: "4/21/2022"
output: html_document
params:
  mrna_experiment_id: "mrna_demo"
  small_rna_experiment_id: "small_rna_demo"
  integrative_experiment_id: "integrative_demo"
  
  mrna_software_method: "rsem"
  small_rna_software_method: "tiny_rna"
  
  mrna_counts_matrix: ""
  small_rna_counts_matrix: "small_rna_counts_matrix.csv"
  
  mrna_metadata: "demo_mrna_metadata.csv"
  small_rna_metadata: "demo_small_rna_metadata.csv"
  
  mrna_gene_table_method: "full_table"
  small_rna_gene_table_method: "full_table"
  
  gene_table: "demo_gene_table.csv"
  
  generate_results_tables: TRUE
  
  generate_pca: TRUE
  
  generate_intra_condition: TRUE
  
  generate_mean_reads: TRUE
  save_mean_reads_interactive: FALSE
  p_value_threshold: 0.05
  fold_change_threshold: 1.3
  lower_transparency: 0.2
  upper_transparency: 0.9
  customize_by_class: TRUE
  customize_by_significance: FALSE
  mrna_class_parameters: "demo_class_parameters.csv"
  small_rna_class_parameters: "demo_class_parameters.csv"
  
  generate_ma: TRUE
  
  generate_heatmap: TRUE
  heatmap_type: "all_classes"
  heatmap_selected_classes: ""
  
  cross_comparisons: ""
  
  generate_integrative_results_tables: TRUE
  
  generate_cosmic_plots: TRUE
  
  generate_slope_plots: TRUE
  
  slope_plot_classes: "ALG,CSR,miRNA,WAGO"
---

```{r params1,echo=FALSE,include=FALSE,warning=FALSE}

# Install Knitr if Necessary
if (require(knitr) == FALSE) {
  install.packages("knitr")
}

# Import Knitr
library(knitr)

# Establish Small RNA Parameters
small_rna_params = list(experiment_id = params$small_rna_experiment_id,
                        software_method = params$small_rna_software_method,
                        counts_matrix = params$small_rna_counts_matrix,
                        metadata = params$small_rna_metadata,
                        gene_table_method = params$small_rna_gene_table_method,
                        gene_table = params$gene_table,
                        generate_results_tables = params$generate_results_tables,
                        generate_pca = params$generate_pca,
                        generate_intra_condition = params$generate_intra_condition,
                        generate_mean_reads = params$generate_mean_reads,
                        save_mean_reads_interactive = params$save_mean_reads_interactive,
                        p_value_threshold = params$p_value_threshold,
                        fold_change_threshold = params$fold_change_threshold,
                        lower_transparency = params$lower_transparency,
                        upper_transparency = params$upper_transparency,
                        customize_by_class = params$customize_by_class,
                        customize_by_significance = params$customize_by_significance,
                        class_parameters = params$small_rna_class_parameters,
                        generate_ma = params$generate_ma,
                        generate_heatmap = params$generate_heatmap,
                        heatmap_type = params$heatmap_type,
                        heatmap_selected_classes = params$heatmap_selected_classes)

# Establish mRNA Parameters
mrna_params = list(experiment_id = params$mrna_experiment_id,
                   software_method = params$mrna_software_method,
                   counts_matrix = params$mrna_counts_matrix,
                   metadata = params$mrna_metadata,
                   gene_table_method = params$mrna_gene_table_method,
                   gene_table = params$gene_table,
                   generate_results_tables = params$generate_results_tables,
                   generate_pca = params$generate_pca,
                   generate_intra_condition = params$generate_intra_condition,
                   generate_mean_reads = params$generate_mean_reads,
                   save_mean_reads_interactive = params$save_mean_reads_interactive,
                   p_value_threshold = params$p_value_threshold,
                   fold_change_threshold = params$fold_change_threshold,
                   lower_transparency = params$lower_transparency,
                   upper_transparency = params$upper_transparency,
                   customize_by_class = params$customize_by_class,
                   customize_by_significance = params$customize_by_significance,
                   class_parameters = params$mrna_class_parameters,
                   generate_ma = params$generate_ma,
                   generate_heatmap = params$generate_heatmap,
                   heatmap_type = params$heatmap_type,
                   heatmap_selected_classes = params$heatmap_selected_classes)

# Establish Integrative Parameters
integrative_params = list(integrative_experiment_id = params$integrative_experiment_id,
                          gene_table = params$gene_table,
                          cross_comparisons = params$cross_comparisons,
                          generate_integrative_results_tables = params$generate_integrative_results_tables,
                          generate_cosmic_plots = params$generate_cosmic_plots,
                          generate_slope_plots = params$generate_slope_plots,
                          slope_plot_classes = params$slope_plot_classes)

# Remove Parameters
rm(params)

```

```{r small_rna,include=FALSE}

# Execute Small RNA Pipeline
rmarkdown::render("Montgomery_DESeq2_Pipeline.Rmd",output_file = paste0(format(Sys.Date(),format = "%d_%m_%y_"),small_rna_params$experiment_id,".html"),params = small_rna_params)

# Remove Parameters
rm(params)

# Store Objects for Integrative Analysis
small_rna_dds = dds
small_rna_cts = cts
small_rna_cts_names = cts_names
small_rna_common_names = common_names
small_rna_class_names = class_names
small_rna_res = res
small_rna_combinations = combinations
colnames(small_rna_combinations) = paste0(small_rna_combinations[2,],"_vs_",small_rna_combinations[1,])

```

```{r mrna,include=FALSE}

# Execute mRNA Pipeline
rmarkdown::render("Montgomery_DESeq2_Pipeline.Rmd",output_file = paste0(format(Sys.Date(),format = "%d_%m_%y_"),mrna_params$experiment_id,".html"),params = mrna_params)

# Remove Parameters
rm(params)

# Store Objects for Integrative Analysis
mrna_dds = dds
mrna_cts = cts
mrna_cts_names = cts_names
mrna_common_names = common_names
mrna_class_names = class_names
mrna_res = res
mrna_combinations = combinations
colnames(mrna_combinations) = paste0(mrna_combinations[2,],"_vs_",mrna_combinations[1,])

```

```{r integrative-setup,include=FALSE}

# Install (if Necessary) and Load Plotly Package
if (require(plotly) == FALSE) {
  install.packages("plotly")
}
library(plotly)

# Create Output Directory
dir.create(paste0(getwd(),"/",format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_output/"))

# Save YAML Parameters
yaml::write_yaml(integrative_params,paste0(format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_output/",
                                           integrative_params$integrative_experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"params.yml"))

# Import Gene Table
gene_table = read.csv(integrative_params$gene_table)

# Generate Matrix of Small RNA and mRNA Target Combinations
target_matrix = NULL
for (i in 1:length(gene_table[,1])) {
  if (gene_table[i,1] %in% names(small_rna_cts_names)) {
    for (j in strsplit(gene_table[i,4],",")[[1]]) {
      if (j %in% names(mrna_cts_names)) {
        target_matrix = rbind(target_matrix,c(gene_table[i,1],j))
      }
    }
  }
}

# Cosmic Plot Function
cosmic_plot = function(target_data,comparison_name) {
  
  # Produce Target Data Subsets
  target_data_s_sig_m_insig = subset(target_data,target_data$small_rna_p_value < 0.05 & target_data$mrna_p_value > 0.05)
  target_data_s_sig_m_sig = subset(target_data,target_data$small_rna_p_value < 0.05 & target_data$mrna_p_value < 0.05)
  target_data_s_insig_m_insig = subset(target_data,target_data$small_rna_p_value > 0.05 & target_data$mrna_p_value > 0.05)
  target_data_s_insig_m_sig = subset(target_data,target_data$small_rna_p_value > 0.05 & target_data$mrna_p_value < 0.05)
  
  # Plotly Call
  cosmic_plot = plot_ly(width = 1250,height = 850) %>% 
    
    # Plot Points Significant for Both Small RNA and mRNA
    add_trace(data = target_data_s_sig_m_sig,type = "scatter",mode = "markers",
              x = ~small_rna_fold_change,y = ~mrna_fold_change,
              size = ~mrna_mean,sizes = c(5,500),
              color = ~log2(small_rna_mean),
              opacity = 1.0,
              marker = list(line = list(color = "#808080",width = 1)),
              colors = c("#C6DBEF","#9ECAE1","#6BAED6","#4292C6","#2171B5",
                         "#08519C","#08306B","#041b3c","#000000"),
              text = ~paste0("<b>small rna p < 0.05<br>mrna p < 0.05</b><br>",
                             "<br>small rna: ",small_rna_cts_names[small_rna],
                             "<br>small rna class: ",small_rna_class,"<br>",
                             "<br>mrna target: ",mrna_cts_names[mrna_target],
                             "<br>mrna class: ",mrna_class),
              hoverinfo = "text",
              name = "small rna p < 0.05<br>mrna p < 0.05") %>% 
    
    # Plot Points Significant for Small RNA, Insignificant for mRNA
    add_trace(data = target_data_s_sig_m_insig,type = "scatter",mode = "markers",
              x = ~small_rna_fold_change,y = ~mrna_fold_change,
              size = I(3),
              color = ~log2(small_rna_mean),
              opacity = 0.6,
              colors = c("#C6DBEF","#9ECAE1","#6BAED6","#4292C6","#2171B5",
                         "#08519C","#08306B","#041b3c","#000000"),
              text = ~paste0("<b>small rna p < 0.05<br>mrna p > 0.05</b><br>",
                             "<br>small rna: ",small_rna_cts_names[small_rna],
                             "<br>small rna class: ",small_rna_class,"<br>",
                             "<br>mrna target: ",mrna_cts_names[mrna_target],
                             "<br>mrna class: ",mrna_class),
              hoverinfo = "text",
              name = "small rna p < 0.05<br>mrna p > 0.05") %>% 
    
    # Plot Points Insignificant for Small RNA, Significant for mRNA
    add_trace(data = target_data_s_insig_m_sig,type = "scatter",mode = "markers",
              x = ~small_rna_fold_change,y = ~mrna_fold_change,
              size = ~mrna_mean,sizes = c(5,500),
              color = I("#808080"),
              marker = list(line = list(color = "#000000",width = 1)),
              opacity = 0.2,
              text = ~paste0("<b>small rna p > 0.05<br>mrna p < 0.05</b><br>",
                             "<br>small rna: ",small_rna_cts_names[small_rna],
                             "<br>small rna class: ",small_rna_class,"<br>",
                             "<br>mrna target: ",mrna_cts_names[mrna_target],
                             "<br>mrna class: ",mrna_class),
              hoverinfo = "text",
              name = "small rna p > 0.05<br>mrna p < 0.05") %>% 
    
    # Plot Points Insignificant for Both Small RNA and mRNA
    add_trace(data = target_data_s_insig_m_insig,type = "scatter",mode = "markers",
              x = ~small_rna_fold_change,y = ~mrna_fold_change,
              size = I(3),
              color = I("#808080"),
              opacity = 0.2,
              text = ~paste0("<b>small rna p > 0.05<br>mrna p > 0.05</b><br>",
                             "<br>small rna: ",small_rna_cts_names[small_rna],
                             "<br>small rna class: ",small_rna_class,"<br>",
                             "<br>mrna target: ",mrna_cts_names[mrna_target],
                             "<br>mrna class: ",mrna_class),
              hoverinfo = "text",
              name = "small rna p > 0.05<br>mrna p > 0.05") %>%
    
    # Specify Plot Layout, Legend, and Annotations
    layout(title = list(text = paste0("<b>",comparison_name,"</b>"),font = list(size = 24),
                        xref = "paper",x = 0,xanchor = "left"),
           xaxis = list(title = "small rna log2 fold change"),
           yaxis = list(title = "mrna log2 fold change"),
           margin = list(t = 100),
           legend = list(title = list(text = "<b>statistical significance</b>"),y = 1),
           annotations = list(text = "<b>point size</b>: mrna mean<br>(for significant mrna genes)",
                              font = list(size = 14),
                              align = "left",
                              xref = "paper",x = 1.02,xanchor = "left",yref = "paper",y=0.75,
                              showarrow = FALSE)) %>% 
    
    # Specify Colorbar Layout
    colorbar(title = "<b>color</b>: small rna log2 mean<br>(for significant small rna genes)<br> <br>",
             x = 1.02,xpad = 0,xanchor = "left",y=0.7)
  
  # Object Return
  return(cosmic_plot)
}

# Slope Plot Function
slope_plot = function(target_data,comparison_name,slope_plot_classes) {
  
  # Build Slope Plot Data Matrix From Target Data
  slope_plot_data = rbind(cbind(rep("small_rna",dim(target_data)[1]),1:dim(target_data)[1],as.matrix(target_data[,1:5])),
                          cbind(rep("mrna",dim(target_data)[1]),1:dim(target_data)[1],as.matrix(target_data[,6:10])))
  
  # Rename Columns
  colnames(slope_plot_data) = c("type","index","gene_id","class","mean","fold_change","p_value")
  
  # Convert Slope Plot Data to Data Frame
  slope_plot_data = data.frame(slope_plot_data)
  
  # Isolate Data to Include Only Listed Slope Plot Classes
  slope_plot_data = slope_plot_data[which(slope_plot_data$index %in% which(slope_plot_data$class %in% slope_plot_classes)),]
  slope_plot_data$class = target_data$small_rna_class[as.numeric(slope_plot_data$index)]
  
  # Eliminate Listed Slope Plot Classes Not Present in the Data
  slope_plot_classes = slope_plot_classes[which(slope_plot_classes %in% slope_plot_data$class)]
  
  # Default Colors List
  class_colors = c("#1F78B4","#1B9E77","#E6AB02","#E7298A","#E31A1C","#6A3D9A","#A6761D","#000000",
                   "#33A02C","#014421","#FF7F00","#8B0000","#FF9E84","#907567","#525252")
  
  # Assign Color Values by Iterating Through Default List
  color_values = rep(class_colors,ceiling(length(slope_plot_classes)/length(class_colors)))[1:length(slope_plot_classes)]
  names(color_values) = sort(slope_plot_classes)
  
  # Assign Transparency Values
  alpha_values = NULL
  for (i in sort(slope_plot_classes)) {
    
    # Assign Value Inversely Proportional to Number of Genes in Each Class
    alpha_values = c(alpha_values,1/(1+(length(which(slope_plot_data$class == i))/100)))
    
  }
  
  # GGplot2 Call
  slope_plot = ggplot(data = slope_plot_data,
                    aes(x = factor(type,levels = c("small_rna","mrna")),y = as.numeric(fold_change),group = index)) + 
    geom_line(aes(color = factor(class),alpha = factor(class)),size = 0.5) + 
    scale_color_manual(values = color_values,name = "Feature Class") + 
    scale_alpha_manual(values = alpha_values,guide = "none") + 
    xlab("Feature Type") + 
    scale_x_discrete(expand = expansion(mult = 0)) + 
    ylab("Log2 Fold Change") + 
    ggtitle(comparison_name) + 
    theme_classic() + 
    theme(axis.text = element_text(size = 12),axis.title = element_text(size = 12)) 
  
  # Object Return
  return(slope_plot)
}

```

```{r integrative-render,include=FALSE}

# Knit Object Setup
integrative_results_tables_out = NULL
cosmic_plots_out = NULL
slope_plots_out = NULL

# Pass message to HTML output if plots are not marked for rendering
if (integrative_params$generate_integrative_results_tables == FALSE) {
  integrative_results_tables_out = "Integrative Results Tables not marked for rendering"
}
if (integrative_params$generate_cosmic_plots == FALSE) {
  cosmic_plots_out = "Cosmic Plots not marked for rendering"
}
if (integrative_params$generate_slope_plots == FALSE) {
  slope_plots_out = "Slope Plots not marked for rendering"
}

# Produce Combination List for Interactive Tables and Plots
if (integrative_params$cross_comparisons == "") {
  cross_comparisons = intersect(colnames(small_rna_combinations),colnames(mrna_combinations))
} else {
  cross_comparisons = unlist(strsplit(integrative_params$cross_comparisons,","))
}

# Build List of Slope Plot Classes
if (integrative_params$slope_plot_classes == "") {
  slope_plot_classes = levels(as.factor(gene_table[,3]))
} else {
  slope_plot_classes = unlist(strsplit(integrative_params$slope_plot_classes,","))
}

# Comparison-Wise Rendering
for (i in 1:length(cross_comparisons)) {
  
  # Match Intended Comparisons with Results Object IDs
  small_rna_res_id = which(colnames(small_rna_combinations) == cross_comparisons[i])
  mrna_res_id = which(colnames(mrna_combinations) == cross_comparisons[i])
  
  # Construct Target Data Matrix
  target_data = na.omit(as.matrix(cbind(target_matrix[,1],small_rna_class_names[target_matrix[,1]],
                                        small_rna_res[[small_rna_res_id]][target_matrix[,1],c(1,2,6)],
                                        target_matrix[,2],mrna_class_names[target_matrix[,2]],
                                        mrna_res[[mrna_res_id]][target_matrix[,2],c(1,2,6)])))
  
  # Define Target Data Matrix Column Names
  colnames(target_data) = c("small_rna","small_rna_class",
                            "small_rna_mean","small_rna_fold_change","small_rna_p_value",
                            "mrna_target","mrna_class",
                            "mrna_mean","mrna_fold_change","mrna_p_value")
  
  # Convert Appropriate Columns to Numeric Data
  target_data = data.frame(target_data)
  target_data$small_rna_mean = as.numeric(target_data$small_rna_mean)
  target_data$small_rna_fold_change = as.numeric(target_data$small_rna_fold_change)
  target_data$small_rna_p_value = as.numeric(target_data$small_rna_p_value)
  target_data$mrna_mean = as.numeric(target_data$mrna_mean)
  target_data$mrna_fold_change = as.numeric(target_data$mrna_fold_change)
  target_data$mrna_p_value = as.numeric(target_data$mrna_p_value)
  
  # Assign Object for Knitting Plots
  assign(paste0("target_data",i,"_plots"),target_data)
  
  # Integrative Results Tables Rendering and Saving
  if (integrative_params$generate_integrative_results_tables == TRUE) {
    
    # Round Numeric Columns
    for (j in c(3:5,8:10)) {
      target_data[,j] = round(as.numeric(target_data[,j]),digits = 3)
    }
    
    # Assign Object for Knitting Tables
    assign(paste0("target_data",i),target_data)
    
    # Table Saving
    write.csv(target_data,paste0(format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_output/",
                                 format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_",
                                 cross_comparisons[i],"_integrative_data.csv"),row.names = FALSE)
    
    # Knit Object Generation
    integrative_results_tables_out = c(integrative_results_tables_out,knit_expand(text = paste0("### ",sub("_vs_"," vs ",cross_comparisons[i]))))
    integrative_results_tables_out = c(integrative_results_tables_out,knit_expand(text = "```{r integrative-tables{{i}},echo=FALSE,warning=FALSE}\n datatable(target_data{{i}})\n```\n"))
    
  } else {
    
    # Null Plots Rendering
    integrative_results_tables_out = c(integrative_results_tables_out,"")
    
  }
  
  # Cosmic Plot Rendering and Saving
  if (integrative_params$generate_cosmic_plots == TRUE) {
    
    # HTML Widget Saving
    htmlwidgets::saveWidget(cosmic_plot(target_data,sub("_vs_"," vs ",cross_comparisons[i])),
                            paste0(format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_output/",
                                   format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_",
                                   cross_comparisons[i],"_cosmic_plot.html"))
    
    # Remove Accessory Cosmic Plot Files
    unlink(paste0(format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_output/",
                  format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_",
                  cross_comparisons[i],"_cosmic_plot_files/"),recursive = TRUE)
    
    
    # Knit Object Generation
    cosmic_plots_out = c(cosmic_plots_out,knit_expand(text = paste0("### ",sub("_vs_"," vs ",cross_comparisons[i]))))
    cosmic_plots_out = c(cosmic_plots_out,knit_expand(text = "```{r cosmic-plots{{i}},echo=FALSE,warning=FALSE}\n cosmic_plot(target_data{{i}}_plots,sub('_vs_',' vs ',cross_comparisons[{{i}}]))\n```\n"))
    
  } else {
    
    # Null Plots Rendering
    cosmic_plots_out = c(cosmic_plots_out,"")
    
  }
  
  # Slope Plot Rendering and Saving
  if (integrative_params$generate_slope_plots == TRUE) {
    
    # Plot Saving
    pdf(paste0(format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_output/",
               format(Sys.Date(),format = "%d_%m_%y_"),integrative_params$integrative_experiment_id,"_",
               cross_comparisons[i],"_slope_plot.pdf"),
        title = paste0("Slope Plot"),height = 6.5,width = 6.5)
    print(slope_plot(target_data,sub("_vs_"," vs ",cross_comparisons[i]),slope_plot_classes))
    dev.off()
    
    # Knit Object Generation
    slope_plots_out = c(slope_plots_out,knit_expand(text = paste0("### ",sub("_vs_"," vs ",cross_comparisons[i]))))
    slope_plots_out = c(slope_plots_out,knit_expand(text = "```{r slope-plot{{i}},echo=FALSE,warning=FALSE}\n slope_plot(target_data{{i}}_plots,sub('_vs_',' vs ',cross_comparisons[{{i}}]),slope_plot_classes)\n```\n"))
    
  } else {
    
    # Null Plots Rendering
    slope_plots_out = c(slope_plots_out,"")
    
  }
}

```

Welcome to the Montgomery DESeq2 Pipeline for small RNA and mRNA integrative experiments. This pipeline allows for the analysis and visualization of data describing the interactions between small RNA and mRNA components of the same experiment. First, separate runs of the Montgomery R Markdown pipeline for individual experiments are executed, once for the small RNA data and once for the mRNA data. Then, information found in an imported gene table describing targets for each small RNA or mRNA gene is used to produce integrative results tables, cosmic scatter plots, and slope plots. To see more information about the individual experiment Markdown pipeline, visit https://github.com/MontgomeryLab/DESeq2App/tree/main/Markdown_Pipeline. To see more information about the integrative R Markdown pipeline, visit https://github.com/MontgomeryLab/DESeq2App/tree/main/Integrative_Pipeline.

## Integrative Results Tables

Integrative Results Tables include the class, base mean, fold change, and negative binomial test p-value for every small RNA and mRNA target pairing for which valid data is available (some low count genes have indeterminable fold changes and p-values. These genes have therefore been excluded). Classes are derived from the imported gene table. Base mean, fold change, and p-value entries are derived from the DESeq2 analysis. 

## Rendered Plots {.tabset}

`r paste(knit(text = integrative_results_tables_out),collapse = '\n')`

\newpage

## Cosmic Plots

Cosmic Plots utilize ggplot2 techniques for representing five-dimensions of data within a two-dimensional scatter plot. The log2 fold change of the small RNA gene in each small RNA and mRNA pairing is represented on the x-axis, while the log2 fold change of the mRNA target of the corresponding small RNA is represented on the y-axis. Furthermore, pairings with an insignificant small RNA p-value (p > 0.05) are plotted in light grey, and pairings with a significant mRNA p-value (p < 0.05) are given a dark grey border around the point. Points are sized according to their mRNA log2 mean, and they are colored (if the small RNA p-value is significant) on a modified 'blues' color scale from the RColorBrewer packageaccording to their small RNA mean (not log2 transformed for better texturing), with darker blues and black representing higher means. Thus, a comprehensive view of every small RNA and mRNA pairing across the two individual experiments can be visualized amongst the plot's four quadrants. Using the plotly package, these scatter plots offer zooming, panning, significance group isolation, and hover text features. Plots for each experimental contrast (provided the contrast is present in both experiments and specified in the associated yaml parameter) are saved as html widgets, which can be printed within a browser window to high-quality pdf images. 

## Rendered Plots {.tabset}

`r paste(knit(text = cosmic_plots_out),collapse = '\n')`

\newpage

## Slope Plots

Slope Plots provide another way of visualizing fold change relationships between small RNA and mRNA pairings, namely by tracing a straight line between the fold change level of the small RNA gene on the left side of the plot and the fold change level of the corresponding mRNA target on the right side of the plot. Lines are colored by their class as specified in the gene table (colors are built from a high-contrast, accessible palette developed by the Montgomery Lab), and it is recommended that classes representing smaller numbers of genes be singled out for plotting using the associated yaml parameter. This plot may be most useful for analyzing micro RNA (miRNA) and mRNA target relationships from one experimental condition to the other. Furthermore, line opacity for a certain class is inversely proportional to the number of genes in that class according to the function opacity = (1/(1 + n/100)), where n is the number of genes for a class. Thus, classes with an abundance of genes will appear more transparent. Lines from classes with over 5000 genes will be colored grey automatically in order to preserve visualization texture. Plots for each experimental contrast (provided the contrast is present in both experiments and specified in the associated yaml parameter) are saved as pdf files. 

## Rendered Plots {.tabset}

`r paste(knit(text = slope_plots_out),collapse = '\n')`

\newpage
