---
title: "Montgomery DESeq2 Pipeline"
author: "Dr. Taiowa Montgomery and Spencer Kuhn"
date: "3/3/2022"
output: html_document
params:
  experiment_id: "worm1"
  software_method: "Counts Matrix"
  counts_matrix: "count_matrix.csv"
  metadata: "metadata.csv"
  gene_table_method: "Full Table"
  gene_table: "gene_table.csv"
  generate_results_tables: TRUE
  generate_pca: TRUE
  generate_intra_condition: TRUE
  generate_mean_reads: TRUE
  save_mean_reads_interactive: FALSE
  p_value_threshold: 0.05
  fold_change_threshold: 1.3
  lower_transparency: 0.2
  upper_transparency: 0.9
  customize_by_class: TRUE
  customize_by_significance: FALSE
  class_parameters: "class_parameters.csv"
  generate_ma: TRUE
  generate_heatmap: TRUE
  heatmap_type: "Complete"
  heatmap_selected_classes: "ALG,CSR,WAGO"
---

```{r,echo=FALSE,include=FALSE,warning=FALSE}

# Package Installation (if Necessary)
if (require(knitr) == FALSE) {
  install.packages("knitr")
}
if (require(BiocManager) == FALSE) {
  install.packages("BiocManager")
}
if (require(tximport) == FALSE) {
  BiocManager::install("tximport",force = TRUE,update = FALSE)
}
if (require(DESeq2) == FALSE) {
  BiocManager::install("DESeq2",force = TRUE,update = FALSE)
}
if (require(DT) == FALSE) {
  install.packages("DT")
}
if (require(ggplot2) == FALSE) {
  install.packages("ggplot2")
}
if (require(heatmaply) == FALSE) {
  install.packages("heatmaply")
}

# Environment Setup
library(knitr)
library(tximport)
library(DESeq2)
library(DT)
library(ggplot2)
library(heatmaply)

# Create Output Directory
dir.create(paste0(getwd(),"/",format(Sys.Date(),format = "%d_%m_%y_"),params$experiment_id,"_output/"))

# Save YAML Parameters
yaml::write_yaml(params,paste0(format(Sys.Date(),format = "%d_%m_%y_"),params$experiment_id,"_output/",
                               params$experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"params.yml"))

```

```{r,include=FALSE,results='hide',message=FALSE}

# Tabulation Software, Counts Files, and Metadata

# Conditional Import of Counts Matrix
if (params$software_method == "Counts Matrix") {counts_matrix = read.csv(params$counts_matrix)} else {counts_matrix = NULL}

# Conditional Import of tiny RNA Matrix
if (params$software_method == "tiny RNA") {tiny_rna_matrix = read.csv(params$counts_matrix)} else {tiny_rna_matrix = NULL}

# Import Metadata File
metadata = read.csv(params$metadata)

# Import Experiment ID
experiment_id = params$experiment_id

# Repeated Values Solver for Gene Table and Count Matrices/tiny RNA Matrices
repeated_values_solver = function(ids) {
  
  # Check for Repeated Values, Then Add Subscripts
  if (length(ids) == length(unique(ids))) {
    return(ids)
    
  } else {
    
    # Produce List of Subscripts
    ids_reps = rep(NA,length(ids))
    for (i in 1:length(ids)) {
      if (length(which(ids[i] == ids[1:i])) > 1) {
        ids_reps[i] = paste0("_",as.character(length(which(ids[i] == ids[1:i]))))
      } else {
        ids_reps[i] = ""
      }
    }
    
    # Add Subscripts to Repeated Values
    for (i in 1:length(ids)) {
      ids[i] = paste0(ids[i],ids_reps[i])
    }
    
    # Object Return
    return(ids)
    
  }
}

# Counts Matrix Preparation
counts_matrix_prep = function(counts_matrix) {
  
  # Reconcile Repeated Values
  counts_matrix[,1] = repeated_values_solver(counts_matrix[,1])
  
  # Convert First Column to Row Names
  rownames(counts_matrix) = counts_matrix[,1]
  counts_matrix = counts_matrix[,-1]
  
  # Round Numeric Values
  counts_matrix = round(counts_matrix)
  
  # Convert Numeric Values to Integers
  for (i in 1:length(counts_matrix[1,])) {
    counts_matrix[,i] = as.integer(counts_matrix[,i])
  }
  
  # Object Return
  return(counts_matrix)
  
}

# Tiny RNA Matrix Preparation
tiny_rna_matrix_prep = function(tiny_rna_matrix) {
  
  # Separate Common Names
  tiny_rna_matrix[,1] = tiny_rna_matrix[,2]
  for (i in 1:length(tiny_rna_matrix[,1])) {
    if (!is.na(strsplit(tiny_rna_matrix[i,1],", ")[[1]][2])) {
      tiny_rna_matrix[i,2] = strsplit(tiny_rna_matrix[i,1],", ")[[1]][2]
    } else {
      tiny_rna_matrix[i,2] = strsplit(tiny_rna_matrix[i,1],", ")[[1]][1]
    }
    tiny_rna_matrix[i,1] = strsplit(tiny_rna_matrix[i,1],", ")[[1]][1]
  }
  
  # Reconcile Repeated Values
  tiny_rna_matrix[,1] = repeated_values_solver(tiny_rna_matrix[,1])
  tiny_rna_matrix[,2] = repeated_values_solver(tiny_rna_matrix[,2])
  
  # Convert First Column to Row Names
  rownames(tiny_rna_matrix) = tiny_rna_matrix[,1]
  tiny_rna_matrix = tiny_rna_matrix[,-1]
  
  # Round Numeric Values
  tiny_rna_matrix[,-c(1,2)] = round(tiny_rna_matrix[,-c(1,2)])
  
  # Convert Numeric Values to Integers
  for (i in 3:length(tiny_rna_matrix[1,])) {
    tiny_rna_matrix[,i] = as.integer(tiny_rna_matrix[,i])
  }
  
  # Object Return
  return(tiny_rna_matrix)
  
}

# DESeq2 Initialization

# Sample Table Preparation
sample_table = data.frame(condition = as.factor(metadata[,3]))
rownames(sample_table) = metadata[,2]

# Conditional DESeq Data Import Calls
if (params$software_method == "HTSeq") {
  
  # HTSeq Option
  metadata_htseq = data.frame(sampleName = metadata[,2],fileName = metadata[,1],condition = factor(metadata[,3]))
  dds = DESeqDataSetFromHTSeqCount(metadata_htseq,getwd(),~condition)
  
} else if (params$software_method == "Counts Matrix") {
  
  # Counts Matrix/FeatureCounts Option
  counts_matrix = counts_matrix_prep(counts_matrix)
  colnames(counts_matrix) = rownames(sample_table)
  dds = DESeqDataSetFromMatrix(counts_matrix,sample_table,~condition)
  
} else if (params$software_method == "tiny RNA") {
  
  # tiny RNA Option
  tiny_rna_matrix = tiny_rna_matrix_prep(tiny_rna_matrix)[,-c(1,2)]
  colnames(tiny_rna_matrix) = rownames(sample_table)
  dds = DESeqDataSetFromMatrix(tiny_rna_matrix,sample_table,~condition)
  
} else {
  
  # RSEM, Salmon, and Kallisto Options
  txi = tximport(metadata[,1],type = params$software_method,txIn = FALSE,txOut = FALSE)
  txi$length[txi$length == 0] = 1
  dds = DESeqDataSetFromTximport(txi,sample_table,~condition)
  
}

# DESeq2 Analysis
dds = DESeq(dds)

# Normalized Counts Table Assembly
cts = counts(dds,normalized = TRUE)

# Contrast List Assembly Function
combination_generator = function() {
  
  # Control Condition Status Preparation
  conditions_set = as.factor(metadata[,3])
  control_status = as.logical(metadata[,4])
  names(control_status) = conditions_set
  control_status = control_status[unique(names(control_status))]
  control_group = names(which(control_status == TRUE))
  
  # Combination Generation
  if (all(!control_status) | sum(control_status > 1)) {
    combinations = combn(levels(conditions_set),2)
  } else {
    combinations = rbind(rep(control_group,length(levels(conditions_set))-1),
                         levels(conditions_set)[-which(levels(conditions_set) == control_group)])
  }
}

# Contrast List Assembly
combinations = combination_generator()

# List of Results Objects by Contrast
res = list()
for (i in 1:length(combinations[1,])) {
  res = append(res,list(results(dds,contrast=c("condition",combinations[2,i],combinations[1,i]))))
}

```

```{r,include=FALSE,message=FALSE}

# Conditional Gene Table Import
if (params$software_method == "tiny RNA") {
  gene_table = tiny_rna_matrix_prep(tiny_rna_matrix)[,c(1,2)]
} else if (params$gene_table_method == "No Table") {
  gene_table = NULL
} else {
  gene_table = read.csv(params$gene_table)
}

# Gene Table Values Preparation
if (is.null(gene_table)) {
  
  # No Gene Table Option
  common_names = NULL
  class_names = NULL
  
} else {
  if (params$gene_table_method == "Full Table" | params$software_method == "tiny RNA") {
    
    # Full Gene Table Option
    common_names = repeated_values_solver(gene_table[,2])
    names(common_names) = repeated_values_solver(gene_table[,1])
    class_names = gene_table[,3]
    names(class_names) = repeated_values_solver(gene_table[,1])
  
  } else if (params$gene_table_method == "Common Names Only") {
    
    # Common Names Only Option
    common_names = repeated_values_solver(gene_table[,2])
    names(common_names) = repeated_values_solver(gene_table[,1])
    class_names = NULL
    
  } else if (params$gene_table_method == "Gene Class Only") {
    
    # Gene Class Only Option
    common_names = NULL
    class_names = gene_table[,2]
    names(class_names) = repeated_values_solver(gene_table[,1])
  }
}

# Replace Gene IDs with Common Names (if Applicable)
cts_names = rownames(cts)
if (!is.null(common_names)) {
  for (i in 1:length(cts_names)) {
    if (cts_names[i] %in% names(common_names)) {
      cts_names[i] = common_names[cts_names[i]]
    }
  }
}

# Reconcile Repeated Values
cts_names = repeated_values_solver(cts_names)

# Assign Gene IDs to Common Names
names(cts_names) = rownames(cts)

```

\newpage

## Results Tables

Results tables display the gene-wise counts from each replicate in a given contrast, 
followed by the fold change value of each gene and the associated p-value (adjusted) of the negative binomial hypothesis test conducted by DESeq2. 
Lower p-values indicate a lower probability of the null hypothesis that counts between the two conditions are derived from the same distributional parameters. 
Furthermore, common gene names from any uploaded gene names table will appear in a column alongside standard gene IDs for the genes that aren't associated with a common name. 
In addition to pdf outputs, these tables can be saved as html widgets with <strong>sorting, searching, and page size customization</strong> features.

```{r,include=FALSE,message=FALSE}

# Base Table Function
results_table = function(res,group1,group2) {
  
  # Results Object Ordering by P-Value
  res_ordered = res[order(res$pvalue),]
  res_ordered = data.frame(res_ordered)
  
  # Isolating Fold Change and Adjusted P-Value Columns
  res_ordered = res_ordered[,c(2,6)]
  
  # Add Counts Columns for the Given Contrast
  res_ordered = cbind(cts[rownames(res_ordered),
                          rownames(sample_table)[which(sample_table == as.character(group1))]],
                      cts[rownames(res_ordered),
                          rownames(sample_table)[which(sample_table == as.character(group2))]],
                      res_ordered)
  
  # Fold Change Column Transformation
  res_ordered$log2FoldChange[which(res_ordered$log2FoldChange > 0)] = 
    2^res_ordered$log2FoldChange[which(res_ordered$log2FoldChange > 0)]
  res_ordered$log2FoldChange[which(res_ordered$log2FoldChange < 0)] = 
    -1/(2^res_ordered$log2FoldChange[which(res_ordered$log2FoldChange < 0)])
  
  # Fold Change Column Renaming
  colnames(res_ordered)[length(colnames(res_ordered))-1] = "Fold_Change"
  
  # Values Rounding to Three Decimal Places
  res_ordered = round(res_ordered,digits = 3)
  
  # Conditional Addition of Class Names Column
  if (!is.null(class_names)) {
    res_ordered = cbind(class_names[rownames(res_ordered)],res_ordered)
    colnames(res_ordered)[1] = "Class"
  }
  
  # Conditional Addition of Common Names Column
  if (!is.null(common_names)) {
    res_ordered = cbind(common_names[rownames(res_ordered)],res_ordered)
    colnames(res_ordered)[1] = "Common_Name"
  }
  
  # Object Return
  return(res_ordered)
  
}

# Table Saving
if (params$generate_results_tables == TRUE) {
  for (i in 1:length(combinations[1,])){
    write.csv(results_table(res[[i]],combinations[1,i],combinations[2,i]),
              paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
                     experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),
                     combinations[2,i],"_vs_",combinations[1,i],"_Results.csv"))
  }
}

```

```{r,include=FALSE}

# Incorporate plots into html output
results_tables_out = NULL

# Pass message to HTML output if plots are not marked for rendering
if (params$generate_results_tables == FALSE) {
  results_tables_out = "Results Tables not marked for rendering"
}

for (i in 1:length(combinations[1,])) {
  if (params$generate_results_tables == TRUE) {
    results_tables_out = c(results_tables_out,knit_expand(text = paste0("### ",combinations[2,i],' vs ',combinations[1,i])))
    results_tables_out = c(results_tables_out,knit_expand(text = "```{r results-tables{{i}},echo=FALSE,warning=FALSE}\n datatable(results_table(res[[{{i}}]],combinations[1,{{i}}],combinations[2,{{i}}]))\n```\n"))
  } else {
    results_tables_out = c(results_tables_out,"")
  }
}

```

## Rendered Tables {.tabset}

`r paste(knit(text = results_tables_out),collapse = '\n')`

\newpage

## PCA Plot

The PCA plot displays loadings of the first two principal components for each sample/biological replicate in the experimental design. 
Colored by experimental condition, the points of the PCA plot provide a visualization of clustering amongst the samples, 
both within conditions and across conditions.

```{r,include=FALSE,message=FALSE}

# Base Plotting Function
pca_input = function() {
  
  # Create PCA Plot
  pca = plotPCA(vst(dds,blind = FALSE),intgroup="condition")
  
  # Add Plot Title
  pca = ggplot_add(ggtitle(paste0(experiment_id," PCA Plot")),pca)
  
  # Revert to Classic Theme
  pca = ggplot_add(theme_classic() + theme(plot.title = element_text(hjust = 0.5)),pca)
  
  # Object Return
  return(pca)
}

# Plot Saving
if (params$generate_pca == TRUE) {
  pdf(paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
             experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"PCA_Plot.pdf"),
      title = paste0(experiment_id," PCA Plot"),
      height = 6.5,width = 8)
  print(pca_input())
  dev.off()
}

```

```{r,echo=FALSE,fig.height=6.5,fig.width=6.5,results='asis'}

# Incorporate Plot Into HTML Output
if (params$generate_pca == TRUE) {
  pca_input()
} else {
  cat("PCA Plot not marked for rendering")
}

```

\newpage

## Intra-Condition Scatterplot

The Intra-Condition Scatter Plots display log2 counts between pairs of biological replicates within each condition of the experimental design. 
Any counts in the data below 1 are replaced with a value of 1 to simplify the log2 transformation. All replicate pairs in each condition are displayed.

```{r,include=FALSE,message=FALSE}

# Base Plotting Function
intra_condition_input = function() {
  
  # Establish List of Conditions
  intra_conditions = levels(as.factor(sample_table[,1]))
  
  # Intra-Condition Combinations Preparation
  intra_combinations = NULL
  for (i in intra_conditions) {
    intra_combinations = cbind(intra_combinations,
                               combn(rownames(sample_table)[which(sample_table[,1] == i)],2))
  }
  
  # Gridded Plots Generation
  par(mfrow = c(length(intra_conditions),ceiling(length(intra_combinations)/(2*length(intra_conditions)))),
      mar = c(3,1,1,1),mgp = c(2,0.5,0),pty = "s")
  for (i in 1:length(intra_combinations[1,])){
    plot(log2(replace(cts,cts < 1,1))[,c(intra_combinations[1,i],intra_combinations[2,i])],
         col="lightskyblue3",pch=15,cex=0.2)
  }
}

# Plot Saving
if (params$generate_intra_condition == TRUE) {
  pdf(paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
             experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"Intra_Condition.pdf"),
      title = paste0(experiment_id," Intra Condition Scatter Plot"),
      height = 6.5,width = 8)
  print(intra_condition_input())
  dev.off()
}

```

```{r,echo=FALSE,fig.height=6.5,fig.width=6.5,results='asis'}

# Incorporate Plot Into HTML Output
if (params$generate_intra_condition == TRUE) {
  intra_condition_input()
} else {
  cat("Intra-Condition Scatter Plot not marked for rendering")
}

```

\newpage

## Mean Reads Scatter Plots

The Mean Reads Scatter Plots display average log2 counts across biological replicates of experimental condition pairs for a provided contrast. 
Average counts with a value of 0 are assigned the value -4 following the log2 transformation. Guidelines are added to assist with visualization. 
Statistically significant genes are less transparent than insignificant genes, and they are colored <strong>blue</strong> in default plots. 
Axes are scaled to more closely resemble the log2 transformation.Unlabeled tick marks, therefore, do not always represent whole number intervals between labeled tick marks. 
In addition to pdf outputs, Mean Reads Scatter Plots can be saved as html widgets with <strong>draggable zooming, panning, and hover text</strong> features. 
To reset the scatter plot axes within the widget, the <strong>home</strong> button in the top right corner of the widget can be pressed.

## Customization

The mean reads scatter plots can be customized by changing the <strong>p-value</strong> and <strong>fold change</strong> thresholds for distinguishing statistically significant genes.
Furthermore, <strong>upper and lower transparency</strong> thresholds can be set for distinguishing statistical significance. The lower threshold corresponds to insignificant genes.
The points of the plot can also be <strong>colored and sized</strong> according to their gene classes as specified by the <strong>gene table</strong>.
If no gene table has been selected/uploaded, only the p-value, fold change, and transparency thresholds will be customizable. If a gene table has been selected/uploaded, a class parameters csv 
can be selected from the working directory. If used, only points/genes corresponding to classes in the table will be plotted. 
If no class parameters csv is selected, every class will be automatically colored from a list of 15 colors, and every point will be sized with a cex value of 0.3. 
Insignificant genes will be colored grey if the <strong>customize by significance</strong> feature is set to <strong>TRUE</strong>.

```{r,include=FALSE,message=FALSE,warning=FALSE}

# Conditional Class Parameters Generation or Import
if (params$customize_by_class == FALSE | params$gene_table_method == "No Table" | params$gene_table_method == "Common Names Only") {
  
  # Conditional Nullification of Class Parameters
  class_parameters = NULL
  
} else {
  if (params$class_parameters == "") {
    
    # Default Colors List
    class_colors = c("#1F78B4","#1B9E77","#33A02C","#014421","#E6AB02","#FF7F00","#FF9E84","#8B0000",
                     "#E31A1C","#E7298A","#6A3D9A","#907567","#A6761D","#525252","#000000")
    
    # Default Table Generation
    class_parameters = data.frame(point_class = levels(as.factor(class_names)),
                                  point_colors = rep(class_colors,ceiling(length(levels(as.factor(class_names)))/length(class_colors)))[1:length(levels(as.factor(class_names)))],
                                  point_sizes = rep(0.3,length(levels(as.factor(class_names)))))
    
    # Default Table Saving
    write.csv(class_parameters,paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
                                      experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"class_parameters.csv"),row.names = FALSE)
    
  } else {
    
    # Custom Table Import
    class_parameters = read.csv(params$class_parameters)
  }
}

# Plot Parameter Values Preparation
if (is.null(class_parameters)) {
  
  # No Plot Parameters Option
  point_color = NULL
  point_size = NULL
  
} else {
  
  # Full Plot Parameters Option
  point_color = class_parameters[,2]
  names(point_color) = class_parameters[,1]
  point_size = class_parameters[,3]
  names(point_size) = class_parameters[,1]
    
}

# Average Counts Table Assembly
cts_avg = NULL
for (i in levels(as.factor(metadata[,3]))) {
  cts_avg = cbind(cts_avg,log2(rowMeans(cts[,metadata[,2][metadata[,3] == i]])))
}

# Replace Infinite Values With -4
cts_avg = replace(cts_avg,cts_avg=='-Inf',-4)

# Isolate Specific Classes
if (!is.null(class_parameters)) {
  cts_avg = cts_avg[intersect(names(class_names[class_names %in% class_parameters[,1]]),names(cts_names)),]
}

# Assign Appropriate Column Names
colnames(cts_avg) = levels(as.factor(metadata[,3]))

# Base Plotting Function
mean_reads_input = function(res,group1,group2) {
  
  # Significant Genes Classification
  res = res[rownames(cts_avg),]
  res_sig = as.data.frame(subset(res,padj < params$p_value_threshold & abs(log2FoldChange) > log2(params$fold_change_threshold)))
  
  # GGplot2 Aesthetic Variables
  x = cts_avg[,group1]
  y = cts_avg[,group2]
  n = cts_names[rownames(cts_avg)]
  
  # Point Significance Variable
  point_sig = rownames(cts_avg) %in% rownames(res_sig)
  
  # Conditional Determination of Size and Color Factor Variables
  
  # Point Colors Assignment
  if (is.null(point_color)) {
    
    # Default Point Colors
    color_class = paste0(c("p < 0.05","p > 0.05")[2-point_sig])
    color_levels = c("#009ACD","#D3D3D3")
    
  } else {
    
    if (params$customize_by_significance == TRUE) {
      
      # Significance-Customized Point Colors
      color_class = paste0(c("p < 0.05","p > 0.05")[2-point_sig])
      color_class[color_class == "p < 0.05"] = class_names[names(n)[color_class == "p < 0.05"]]
      color_levels = c("p > 0.05" = "#D3D3D3",point_color)
      
    } else {
      
      # Fully Customized Point Colors
      color_class = class_names[names(n)]
      color_levels = point_color
      
    }
  }
  
  # Point Sizes Assignment
  if (is.null(point_size)) {
    
    # Default Point Sizes
    size_class = paste0(c("p < 0.05","p > 0.05")[2-point_sig])
    size_levels = c(0.5,0.3)
    
  } else {
    
    # Customized Point Sizes
    size_class = class_names[names(n)]
    size_levels = point_size
    
  }
  
  # Formatting Gene Name Text Variable
  for (i in 1:length(n)) {
    n[i] = paste0("Gene: ",n[i],"<br>","Class: ",class_names[names(n)[i]])
  }
  
  # Final Data Frame Assembly for GGplot
  points_data = data.frame(x,y,color_class,size_class,point_sig,row.names = n)
  
  # Major and Minor Tick Mark Locations
  tick_sep = seq(-4,20,0.8)
  tick_sep = 2^tick_sep
  tick_sep[1:6] = seq(tick_sep[1],tick_sep[6],length.out = 6)
  tick_sep[6:11] = seq(tick_sep[6],tick_sep[11],length.out = 6)
  tick_sep[11:16] = seq(tick_sep[11],tick_sep[16],length.out = 6)
  tick_sep[16:21] = seq(tick_sep[16],tick_sep[21],length.out = 6)
  tick_sep[21:26] = seq(tick_sep[21],tick_sep[26],length.out = 6)
  tick_sep[26:31] = seq(tick_sep[26],tick_sep[31],length.out = 6)
  
  # Tick Mark Labels
  tick_labs = rep("",31)
  tick_labs[seq(1,31,5)] = prettyNum(log2(tick_sep[seq(1,31,5)]),digits=2,format="f")
  
  # GGplott Object Generation
  mean_reads = ggplot(points_data,aes(x,y)) + 
    
    # Point Plotting, Sizing, Coloring, and Text Addition
    geom_point(aes(color = factor(color_class),size = factor(size_class),alpha = factor(point_sig),shape = factor(20),
                   text = paste0(group1,": ",round(x,digits = 3),"<br>",
                                 group2,": ",round(y,digits = 3),"<br>",n))) + 
    
    # Color, Size, and Transparency Levels
    scale_color_manual(values = color_levels,name = "Feature Class") + 
    scale_size_manual(values = size_levels,guide = "none") +
    scale_alpha_manual(values = c(params$lower_transparency,params$upper_transparency),
                       name = "Significance",labels = c("p > 0.05","p < 0.05")) +
    scale_shape(guide = "none") + 
    
    # Guide Lines
    geom_abline(intercept = 0,slope = 1,color = "grey60") + 
    geom_abline(intercept = 1,slope = 1,color = "grey60") +
    geom_abline(intercept = -1,slope = 1,color = "grey60") + 
    geom_hline(yintercept = log2(10),color = "grey60",linetype = "dashed") + 
    geom_vline(xintercept = log2(10),color = "grey60",linetype = "dashed") + 
    
    # Tick Marks
    scale_y_continuous(breaks = log2(tick_sep),labels = tick_labs,
                       limits = c(-4,max(cts_avg[,c(group1,group2)]))) + 
    scale_x_continuous(breaks = log2(tick_sep),labels = tick_labs,
                       limits = c(-4,max(cts_avg[,c(group1,group2)]))) + 
    
    # Title
    ggtitle(paste0(group2,' vs ',group1)) +
    
    # Axis Labeling
    xlab(paste0('Log2 Average reads in ',group1)) +
    ylab(paste0('Log2 Average reads in ',group2)) +
    
    # Theme Specification
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(legend.position = "right")
  
  # Object Return
  return(mean_reads)
}

# Plot Saving
if (params$generate_mean_reads == TRUE) {
  for (i in 1:length(combinations[1,])){
    
    # Save PDF plots if applicable
    mean_reads_plot = mean_reads_input(res[[i]],combinations[1,i],combinations[2,i])
    pdf(paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
               experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),
               combinations[2,i],"_vs_",combinations[1,i],"_Mean_Reads.pdf"),
        title = paste0(experiment_id," ",combinations[2,i]," vs ",combinations[1,i]," Mean Reads Scatter Plot"),
        height = 6.5,width = 8)
    print(mean_reads_plot)
    dev.off()
    
    if (params$save_mean_reads_interactive == TRUE) {
      
      # Save interactive Plots if applicable
      htmlwidgets::saveWidget(ggplotly(mean_reads_plot,tooltip = "text",width = 1000,height = 650),
                              file = paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
                                            experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),
                                            combinations[2,i],"_vs_",combinations[1,i],"_Mean_Reads.html"))
      
      # Remove accessory mean reads files
      unlink(paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
                    experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),
                    combinations[2,i],"_vs_",combinations[1,i],"_Mean_Reads_files/"),recursive = TRUE)
      
    }
  }
}

```

```{r mean-reads-out,include=FALSE}

# Incorporate plots into html output
mean_reads_out = NULL

# Pass message to HTML output if plots are not marked for rendering
if (params$generate_mean_reads == FALSE) {
  mean_reads_out = "Mean Reads Scatter Plots not marked for rendering"
}

for (i in 1:length(combinations[1,])) {
  if (params$generate_mean_reads == TRUE) {
    mean_reads_out = c(mean_reads_out,knit_expand(text = paste0("### ",combinations[2,i],' vs ',combinations[1,i])))
    mean_reads_out = c(mean_reads_out,knit_expand(text = "```{r mean-reads{{i}},echo=FALSE,warning=FALSE,fig.height=6.5,fig.width=8}\n mean_reads_input(res[[{{i}}]],combinations[1,{{i}}],combinations[2,{{i}}]) \n```\n"))
  } else {
    mean_reads_out = c(mean_reads_out,"")
  }
}

```

## Rendered Plots {.tabset}

`r paste(knit(text = mean_reads_out),collapse = '\n')`

\newpage

## MA Plots

The Standard MA Plots are built from the DESeq2 package and display the fold change of a gene over its mean counts value (normalized) for a provided contrast between experimental conditions. 
Genes with a statistically significant p-value (p < 0.05) are colored in <strong>blue</strong>.

```{r,include=FALSE,message=FALSE}

# Base Plotting Function
ma_input = function(res,group1,group2) {
  plotMA(res,ylim=c(-4,4),main = paste0(group2,' vs ',group1))
}

# Plot Saving
if (params$generate_ma == TRUE) {
  for (i in 1:length(combinations[1,])){
    pdf(paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
               experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),
               combinations[2,i],"_vs_",combinations[1,i],"_MA_Plot.pdf"),
        title = paste0(experiment_id," ",combinations[2,i]," vs ",combinations[1,i]," MA Plot"),
        height = 6.5,width = 6.5)
    print(ma_input(res[[i]],combinations[1,i],combinations[2,i]))
    dev.off()
  }
}

```

```{r ma-out,include=FALSE}

# Incorporate plots into html output
ma_out = NULL

# Pass message to HTML output if plots are not marked for rendering
if (params$generate_ma == FALSE) {
  ma_out = "MA Plots not marked for rendering"
}

for (i in 1:length(combinations[1,])) {
  if (params$generate_ma == TRUE) {
    ma_out = c(ma_out,knit_expand(text = paste0("### ",combinations[2,i],' vs ',combinations[1,i])))
    ma_out = c(ma_out,knit_expand(text = "```{r ma{{i}},echo=FALSE,warning=FALSE,fig.height=6.5,fig.width=6.5}\n ma_input(res[[{{i}}]],combinations[1,{{i}}],combinations[2,{{i}}]) \n```\n"))
  } else {
    mean_reads_out = c(mean_reads_out,"")
  }
}

```

## Rendered Plots {.tabset}

`r paste(knit(text = ma_out),collapse = '\n')`

\newpage

## Heatmap

The <strong>Complete</strong> heatmap shows log2 counts across all samples for any genes above a certain mean count threshold of three 
(meaning an average of eight counts across all samples). Using the package <strong>heatmaply</strong>, 
an interactive html widget is rendered with <strong>draggable zooming, panning, and hover text</strong> features. 
To reset the heatmap axes, the home button in the top right corner of the widget can be pressed. Darker blue cells indicate lower log2 counts values,
while darker red cells indicate higher log2 counts values. 
Furthermore, rows (genes) are clustered using the complete hierarchical clustering method in R (hclust) with eudclidean distances.

The <strong>Class-Separated: All Classes</strong> heatmap relies on the gene table to map all genes with an associated class. 
Such genes are grouped alphabetically by class and clustered hierarchically within their respective group. 
Hover text indicates the class assigned to each gene.

The <strong>Class-Separated: Selected Classes</strong> heatmap produces subplots for each class listed in the associated yaml parameter entry.
Heatmaps for each listed class are scaled according to the limits of the complete heatmap, 
although cell sizes are scaled according to the number of genes associated with each class. 

```{r,include=FALSE,message=FALSE}

# Heatmap Type Import
heatmap_type = params$heatmap_type

# Conditional Separation of Heatmap Classes String
if (heatmap_type == "Class-Separated: Selected Classes") {
  heatmap_selected_classes = strsplit(params$heatmap_selected_classes,",")[[1]]
}

# Conditional Reversion of Heatmap Type to "Complete"
if (params$gene_table_method == "No Table" | params$gene_table_method == "Common Names Only") {heatmap_type = "Complete"}

# Heatmap Matrix Generation
cts_heat = log2(cts)[which(rowMeans(log2(cts)) >= 3),]

# Complete Heatmap Base Plotting Function
complete_heatmap = function() {
  
  # Common Names Conversion
  rownames(cts_heat) = cts_names[rownames(cts_heat)]
    
  # Clustering and Re-Ordering
  d = dist(cts_heat,method = "euclidean")
  h = hclust(d,method = "complete")
  cts_heat = cts_heat[h$order,]
    
  # Heatmaply Call
  heatmap_full = heatmaply(cts_heat,
                           colors = c("#020532","#03095B","#053061","#2166AC","#4393C3","#92C5DE","#D1E5F0",
                                      "#F7F7F7","#FDDBC7","#F4A582","#D6604D","#B2182B","#67001F","#000000"),
                           limits = c(floor(min(cts_heat)),ceiling(max(cts_heat))),
                           showticklabels = c(TRUE,FALSE),Colv = FALSE,Rowv = FALSE,
                           main = "Complete Heatmap")
  
  # Object Return
  return(heatmap_full)
  
}

# All Classes Heatmap Base Plotting Function
all_classes_heatmap = function() {
    
  # Class-separated Matrix and Class Text Matrix Setup
  cts_heat_full = NULL
  cts_heat_text = NULL
    
  # Iteration for Each Class
  for (i in 1:length(levels(as.factor(class_names)))) {
    
    # Isolation of Class-Based Matrix
    cts_heat_class = cts_heat[intersect(names(class_names[which(class_names == levels(as.factor(class_names))[i])]),rownames(cts_heat)),]
      
    # Check for Null Matrices Before Integration Into Heatmap
    if (length(cts_heat_class) > 0) {
      
      # Produce Named Class-Based Matrix
      cts_heat_class = matrix(cts_heat_class,nrow = length(cts_heat_class)/length(cts_heat[1,]),byrow = FALSE,
                              dimnames = list(rownames(cts_heat_class),colnames(cts_heat_class)))
        
      # Addition of Class Text Entries to Overall Text Matrix
      cts_heat_text = rbind(cts_heat_text,
                            matrix(rep(paste0("Class: ",class_names[rownames(cts_heat_class)]),dim(cts_heat_class)[2]),
                                   nrow = dim(cts_heat_class)[1],ncol = dim(cts_heat_class)[2]))
        
      # Common Names Conversion
      rownames(cts_heat_class) = cts_names[rownames(cts_heat_class)]
        
      # Clustering and Re-Ordering for Class-Based Matrices of >1 Row
      if (dim(cts_heat_class)[1] > 1) {
        d = dist(cts_heat_class,method = "euclidean")
        h = hclust(d,method = "complete")
        cts_heat_class = cts_heat_class[h$order,]
      }
        
      # Addition of Class-Based Matrix to Overall Matrix
      cts_heat_full = rbind(cts_heat_full,cts_heat_class)
        
    }
  }
    
  # Heatmaply Call
  heatmap_full = heatmaply(cts_heat_full,
                           colors = c("#020532","#03095B","#053061","#2166AC","#4393C3","#92C5DE","#D1E5F0",
                                      "#F7F7F7","#FDDBC7","#F4A582","#D6604D","#B2182B","#67001F","#000000"),
                           limits = c(floor(min(cts_heat)),ceiling(max(cts_heat))),
                           showticklabels = c(TRUE,FALSE),Colv = FALSE,Rowv = FALSE,
                           main = "All Classes Heatmap",margins = c(50,50,50,50),
                           custom_hovertext = cts_heat_text)
  
  # Object Return
  return(heatmap_full)
  
}

# Selected Classes Heatmap Base Plotting Function
selected_classes_heatmap = function() {
  
  heatmap_full = NULL
  
  # Iteration for Each Selected Class
  for (i in 1:length(heatmap_selected_classes)) {
    
    # Isolation of Class-Based Matrix
    cts_heat_class = cts_heat[intersect(names(class_names[which(class_names == heatmap_selected_classes[i])]),rownames(cts_heat)),]
    
    if (length(cts_heat_class) > 0) {
      cts_heat_class = matrix(cts_heat_class,nrow = length(cts_heat_class)/length(cts_heat[1,]),byrow = FALSE,
                              dimnames = list(rownames(cts_heat_class),colnames(cts_heat_class)))
      
      # Common Names Conversion
      rownames(cts_heat_class) = cts_names[rownames(cts_heat_class)]
      
      # Clustering and Re-Ordering for Class-Based Matrices of >1 Row
      if (length(cts_heat_class) > length(cts_heat[1,])) {
        d = dist(cts_heat_class,method = "euclidean")
        h = hclust(d,method = "complete")
        cts_heat_class = cts_heat_class[h$order,]
      }
      
      # Heatmaply Call
      hm_class = heatmaply(cts_heat_class,
                           colors = c("#020532","#03095B","#053061","#2166AC","#4393C3","#92C5DE","#D1E5F0",
                                      "#F7F7F7","#FDDBC7","#F4A582","#D6604D","#B2182B","#67001F","#000000"),
                           limits = c(floor(min(cts_heat)),ceiling(max(cts_heat))),
                           showticklabels = c(TRUE,FALSE),Colv = FALSE,Rowv = FALSE,
                           ylab = heatmap_selected_classes[i],main = "Selected Classes Heatmap",
                           margins = c(50,50,50,50))
      
      # Addition of Heatmaply Object to Subplots List
      heatmap_full = append(heatmap_full,list(hm_class))
    
    }
  }
  
  # Subplot Compilation
  heatmap_full = subplot(heatmap_full,nrows = length(heatmap_selected_classes),shareX = TRUE,titleY = TRUE,margin = 0.005)
  
  # Object Return
  return(heatmap_full)
  
}

# Conditional Heatmap Saving
if (params$generate_heatmap == TRUE) {
  if (heatmap_type == "Complete") {
    
    # Complete Heatmap Plot Saving
    htmlwidgets::saveWidget(complete_heatmap(),file = paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
                                                             experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"Heatmap.html"))
    
  } else if (heatmap_type == "Class-Separated: All Classes") {
    
    # All Classes Heatmap Plot Saving
    htmlwidgets::saveWidget(all_classes_heatmap(),file = paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
                                                                experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"Heatmap.html"))
    
  } else if (heatmap_type == "Class-Separated: Selected Classes") {
    
    # Selected Classes Heatmap Plot Saving
    htmlwidgets::saveWidget(selected_classes_heatmap(),file = paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
                                                                     experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"Heatmap.html"))
    
  }
}

# Remove accessory heatmap files
unlink(paste0(format(Sys.Date(),format = "%d_%m_%y_"),experiment_id,"_output/",
              experiment_id,"_",format(Sys.Date(),format = "%d_%m_%y_"),"Heatmap_files/"),recursive = TRUE)

```

```{r,echo=FALSE,fig.height=6.5,fig.width=8,fig.align="center",results='asis'}

# Incorporate plot into html output
if (params$generate_heatmap == TRUE) {
  if (heatmap_type == "Complete") {
    
    # Complete Heatmap Plot Saving
    complete_heatmap()
    
  } else if (heatmap_type == "Class-Separated: All Classes") {
    
    # All Classes Heatmap Plot Saving
    all_classes_heatmap()
    
  } else if (heatmap_type == "Class-Separated: Selected Classes") {
    
    # Selected Classes Heatmap Plot Saving
    selected_classes_heatmap()
    
  }
} else {
  cat("Heatmap not marked for rendering")
}

```
